\bf Data Assimilation for Modeling and Calibrating of Disease-altered Potassium Channel Kinetics in Mouse Cardiomyocytes

### 2022-09-27 writing/draft.tex
Abstract revised

\begin{abstract}
Activities of cardiac voltage-gated potassium channels (K\textsubscript{v}) are essential in the electrical conduction system of the heart. Even slight changes in K\textsubscript{v} functions can cause fatal heart disease, such as long QT syndrome. Our recent studies found that altered glycosylation causes aberrant electrical signaling in mouse cardiomyocytes and leads to heart failure and early death. Activities of K\textsubscript{v} channels were also affected by altered glycosylation, and K\textsuperscript{+} currents were significantly reduced. Among major cardiac voltage-gated ion channels (VGIC), K\textsubscript{v} channels have a distinctive feature: various isoforms generate their own unique currents contributing to different phases of repolarization of the cardiac action potential (AP). On the other hand, the dominant isoform is responsible for Na\textsuperscript{+} and Ca\textsuperscript{2+} currents. This characteristic raises challenges for studying K\textsubscript{v} channels. Although it is highly desirable to observe the activities of each K\textsubscript{v} isoform separately, only collective outputs can be measured because of their slightly overlapping voltage-dependence of gating. In whole-cell recordings, the sum of K\textsuperscript{+} currents (I\textsubscript{Ksum}) can be measured. Traditional methods apply curve fitting to estimate components of K\textsubscript{v} isoforms from I\textsubscript{Ksum}. Additional curve fitting is used to find kinetics information for the individual K\textsuperscript{+} currents. However, this traditional approach is limited in its ability to determine gating kinetics of K\textsubscript{v} isoforms rigorously. Curve fittings do not account for interactions of multiple gating processes of various isoforms and dynamics according to changes in protocols. Therefore, we propose a novel data assimilation framework for investigating whole-cell K\textsubscript{v} current traces and apply the proposed method to our glycosylation data. First, computer models of K\textsuperscript{+} currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. Second, we develop fractional factorial designs to identify the parameters that have significant impacts on the current shapes. Third, we develop model calibration as nonlinear optimization that couple the \textit{in-silico} K\textsubscript{v} models and \textit{in-vitro} I\textsubscript{K} recordings. Last, low-dimensional embedding is used to visualize distributional difference of kinetic parameters of cells, showing the cellular-level variability. Experimental results show that the data assimilation method can estimate differences in kinetic rate between the control and experiment groups. This method has strong potential to pave a new way to analyze K\textsubscript{v} and heart diseases.
\end{abstract}

>>>>

\begin{abstract}
Potassium channels (K\textsubscript{v}) are critical to the electrical conduction system of the heart, and aberrant activities in K\textsubscript{v} functions can cause fatal heart disease. K\textsubscript{v} are responsible for repolarizing the cardiac action potential (AP), a change of membrane potential over time, representing the net electrical activity in a cardiomyocyte. Repolarization is a complex process that consists of multiple phases to which different K\textsubscript{v} isoforms contribute. For example, in the mouse ventricle, repolarization is roughly delineated as two kinetic components: the rapidly inactivating current I\textsubscript{Kto}, conducted through K\textsubscript{v}4.2, and the slowly inactivating currents such as I\textsubscript{Kslow1} and I\textsubscript{Kslow2}, conducted through K\textsubscript{v}1.5 and K\textsubscript{v}2.1 respectively. It is a unique characteristic of K\textsubscript{v} compared to other cardiac ion channels where dominant isoforms exist. Although it is highly desirable to study individual K\textsubscript{v} isoforms to understand pathological mechanisms in the cardiac conduction system, only the sum of their activities as a whole can be measured in experiments that record whole-cell voltage clamp K\textsubscript{v} currents (I\textsubscript{K}). A data assimilation method, combining experimental data with mathematical/computational models, is used to estimate information on K\textsubscript{v} isoforms. Most existing studies separate individual K\textsuperscript{+} currents mathematically from I\textsubscript{K} by a curve-fitting decomposition procedure, assuming a curve of I\textsubscript{K} consists of the summation of exponential decay functions and a constant term; each represents the shape of K\textsuperscript{+} currents. However, this traditional approach is limited in its ability to scrutinize the kinetics and dynamics of K\textsubscript{v} in two-fold: 1) the curve-fitting I\textsubscript{K} decomposition only relies on the shape of K\textsuperscript{+} current without considering underlying kinetics and interactions of K\textsubscript{v} isoforms, and 2) it handles I\textsubscript{K} traces from the same cell with different conditions one-by-one individually. There is an urgent need to develop new data assimilation methods that simultaneously delineate the underlying kinetics of multiple I\textsubscript{K} traces for a better understanding of the ion channels to provide insights into clinical applications, such as drug design. Therefore, we propose a novel \textit{concurrent data assimilation} framework for investigating the kinetics of K\textsubscript{v} isoforms from multiple I\textsubscript{K} traces measured in the same cell. First, computer models of K\textsuperscript{+} currents are designed with parameters that control the kinetic rates and, in turn, determine the currents. Second, we develop fractional factorial designs to identify the parameters that have significant impacts on the current shapes. Third, we develop model calibration as nonlinear optimization that couple the \textit{in-silico} K\textsubscript{v} models and \textit{in-vitro} I\textsubscript{K} recordings. Last, low-dimensional embedding is used to visualize distributional difference of kinetic parameters of cells, showing the cellular-level variability. We apply the proposed framework to data of glyco-altered I\textsubscript{K} recordings from our previous studies. Experimental results show that the proposed method can estimate differences in kinetics between the control and experiment groups. This proposed method has strong potential to pave a new way to analyze K\textsubscript{v} and heart diseases.
\end{abstract}
